A weakly structured stem for human origins in Africa

### 2023-01-08 paper.tex
Supp fig cite

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single-origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains
prohibitive to fully explore the space of plausible models of early human
population structure, we sought to capture model uncertainty by exploring
multiple parameterizations of early history. The best fit models presented
here include reticulation and migration between early human populations rather
than archaic hominin admixture from long-isolated branches
(Figure~\ref{fig:proposed-models}C). 
Elements of a recent single-origin or African multi-regionalism
(Figure~\ref{fig:proposed-models}A,D)
feature in our best fit models, as indicated in the recent time of
contemporary population divergence
or the gene flow between disparate stems, respectively.
We caution that we cannot rule out that more complex models involving
additional stems, continuous structure, or
hybrid models including both weak structure and archaic hominin admixture may
better explain the data. Because parameters related to the split time,
migration rates, and relative sizes of the early stems were variable across
models, reflecting a degree of confounding among these parameters, we refrained
from introducing additional branches associated with more parameters during
that period.  Rather than interpreting the two stems as representing
well-defined and stable populations over hundreds of thousands of years, we
interpret the weakly structured stem as consistent with a population
coalescence and fragmentation model \citep{Scerri2019-xg}.
Additional African populations such as those 
from Central Africa, other Khoe-San groups or pre-Holocene ancient DNA samples,
could further test our proposed models.
Given that our samples cover major ancestry components across Africa, we predict
that any new admixture signal is likely to be regionally localized.

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S, then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, resulted in the
formation of an ancestral Khoe-San population $\approx120$ka. The second event
$\approx100$ka between Stem 1E and Stem 2, resulted in the formation of the
ancestors of Eastern/Western Africans including the ancestors of ``Out of
Africans''. While $120$ka can be interpreted as the basal human population divergence,
we suggest concpetualizing this event as the time of most recent
shared ancestry among sampled populations.
``Population divergence'' in population genetics is a complex concept because 
of the co-estimation of divergence time and subsequent migration;
methods make differing assumptions regarding the degree of a ``clean'' split 
between descendant groups vs. isolation-with-migration 
(Figure~\ref{fig:supp-IM-misspecification}).
Therefore, wide variation exists in estimates of 
divergence time in the literature (e.g.\citep{Pagani2015-pz}).

@@

Third, multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI
Section~\ref{sec:morpho_diff}). This correspondence potentially allows
predictions of our model to be related to the fossil record. The fossil record
of Africa is sparse during the time period of the stems, but of the fossils
that date to this time period, some are fairly similar in morphology to
contemporary humans (e.g., Omo 1 from Omo Kibish, Ethiopia
\citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological
features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco
\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in
morphology to have been assigned to species other than \emph{Homo sapiens}
(e.g., DH1 from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}). If,
as our model predicts, the genetic differences between the stems were
comparable to those among contemporary human populations, the most
morphologically divergent fossils are unlikely to represent branches that
contributed appreciably to contemporary human ancestries.

>>>>

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single-origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains
prohibitive to fully explore the space of plausible models of early human
population structure, we sought to capture model uncertainty by exploring
multiple parameterizations of early history. The best fit models presented
here include reticulation and migration between early human populations rather
than archaic hominin admixture from long-isolated branches
(Figure~\ref{fig:proposed-models}C). 
Elements of a recent single-origin or African multi-regionalism
(Figure~\ref{fig:proposed-models}A,D)
feature in our best fit models, as indicated in the recent time of
contemporary population divergence
or the gene flow between disparate stems, respectively.

We caution that we cannot rule out that more complex models involving
additional stems, continuous structure, or
hybrid models including both weak structure and archaic hominin admixture may
better explain the data. Because parameters related to the split time,
migration rates, and relative sizes of the early stems were variable across
models, reflecting a degree of confounding among these parameters, we refrained
from introducing additional branches associated with more parameters during
that period.  Rather than interpreting the two stems as representing
well-defined and stable populations over hundreds of thousands of years, we
interpret the weakly structured stem as consistent with a population
coalescence and fragmentation model \citep{Scerri2019-xg}.
Additional African populations such as those 
from Central Africa, other Khoe-San groups or pre-Holocene ancient DNA samples,
could further test our proposed models.
Given that our samples cover major ancestry components across Africa, we predict
that any new admixture signal is likely to be regionally localized.

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S, then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, resulted in the
formation of an ancestral Khoe-San population $\approx120$ka. The second event
$\approx100$ka between Stem 1E and Stem 2, resulted in the formation of the
ancestors of Eastern/Western Africans including the ancestors of ``Out of
Africans''. While $120$ka can be interpreted as the basal human population divergence,
we suggest concpetualizing this event as the time of most recent
shared ancestry among sampled populations.
``Population divergence'' in population genetics is a complex concept because 
of the co-estimation of divergence time and subsequent migration;
methods make differing assumptions regarding the degree of a ``clean'' split 
between descendant groups vs. isolation-with-migration 
(Figure ~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
Therefore, wide variation exists in estimates of 
divergence time in the literature (e.g.\citep{Pagani2015-pz}).

@@

Third, multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI
Section~\ref{sec:morpho_diff}). This correspondence potentially allows
predictions of our model to be related to the fossil record. The fossil record
of Africa is sparse during the time period of the stems, but of the fossils
that date to this time period, some are fairly similar overall in morphology to
contemporary humans (e.g., Omo 1 from Omo Kibish, Ethiopia
\citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological
features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco
\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in
morphology to have been assigned to species other than \emph{Homo sapiens}
(e.g., DH1 from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}). If,
as our model predicts, the genetic differences between the stems were
comparable to those among contemporary human populations, the most
morphologically divergent fossils are unlikely to represent branches that
contributed appreciably to contemporary human ancestries.
